A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization

### 2020-07-13 ms/ms.Rmd
r1

---
title: "A stomatal model of anatomical tradeoffs between photosynthesis and pathogen defense"
runtitle: "Stomata tradeoff photosynthesis for defense"
documentclass: frontiersSCNS
author:
  - name: Christopher D. Muir
    affiliation: '1'
    etal: Muir
    email: cdmuir@hawaii.edu
affiliation:
  - id: '1'
    department: School of Life Sciences
    institution: University of Hawaii
    city: Honolulu
    state: Hawaii
    country: USA
output: 
  rticles::frontiers_article:
    includes:
      in_header: header.tex
csl: frontiers.csl
bibliography: stomata-tradeoff.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Stomatal pores control both leaf gas exchange and are an entry for many plant pathogens, setting up the potential for tradeoffs between photosynthesis and defense. To prevent colonization and limit infection, plants close their stomata after recognizing pathogens. In addition to closing stomata, anatmoical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes would permanently reduce the gas exchange capacity for the life of the leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which determine maximum rates of gas exchange. The model predicts greater stomatal size or density increases colonization, but the effect is most pronounced when stomatal cover is low. I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential conflict between maximizing defense and minimizing stomatal cover. To my knowledge, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. It makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for accurate predictions.

@@

<!-- overview -->
Stomata evolved to regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many plant pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017]. The density and size of stomata set the anatomical maximum rate of stomatal conductance to CO$_2$ and water vapor [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019]. Many plant pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry into the leaf. Since stomatal conductance is a major limitation on photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] while pathogens reduce fitness, this sets up a potential tradeoff between increased photosynthesis and defense against pathogens. Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings. Lack of a theoreatical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange.

<!-- paragraph on stomatal anatomy and gsmax -->
Stomatal anatomy is the key link between gas exchange and pathogen colonization. The density and size of stomata not only determines the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019]. Therefore, I use anatomical \gsmax{} as a proxy for \gsop{} and do not address dynamic changes in stomatal aperture (see [Discussion]). @harrison_influence_2019 recently reviewed the relationship between stomatal anatomy and gas exchange in detail.

<!-- paragraph on stomatal anatomy and pathogen defense -->
Many pathogens rely on stomata to gain entry into the leaf, but it is unclear how anatomical traits that determine \gsmax{} (size and density) affect the ability of pathogens to colonize leaves. The impact of pathogens on host fitness is complex, but after a pathogen reaches a host, the first major step is colonization. Once infected, a pathogen can reduce fitness. Susceptible hosts can lose much of their biomass or die, but even resistant hosts must allocate resources to defense or reduce photosynthesis because of necrosis around sites of infection. Plants can limit colonization physiologically by closing stomata after they recognize pathogens, called stomatal defense [@melotto_stomatal_2017]. Anatomy may be another layer of defense; plants can reduce pathogen colonization by developing leaves with lower stomatal density and/or size. Infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019], which suggests that both anatomical and physiological responses (stomatal closure) affect host colonization. By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area. One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throught the life of the leaf. Because we do not understand the relationship between stomatal anatomy and colonization well, we cannot predict the relationship between \gsmax{} and infection.

<!-- paragraph on size-density scaling? -->
If stomatal size and density affect pathogen colonization, selection to limit colonization may shape stomatal size-density scaling relationships. Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; @brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known. Here I argue that pathogens could shape selection on this relationship. Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or total epidermal area allocated to stomata [\fs; @de_boer_optimal_2016] because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:

@@

$b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material]). If size and density also affect pathogen colonization, then selection from foliar pathogens could significantly alter the size-density scaling relationship. The empirical size-density scaling relationship is linear on a log-log scale, determined by an intercept $\alpha$ and slope $\beta$:

@@

How would adding pathogens alter these predicted scaling relationships? For simplicity, imagine two environments, one without foliar pathogens and one with lots. In the absence of foliar pathogens, we expect size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost combination of them. What happens when we introduce pathogens? Assuming that stomatal size and density increase pathogen colonization, then selection will favor reduced size and/or density. This would change the intercept $\alpha$ but not the slope. The effect of foliar pathogens on the slope depends on the relationship between size, density, and probability of colonization. If the probability of colonization is proportional to the product of *linear* stomatal size ($S ^ {0.5}$) and density ($\propto D S ^ {0.5}$ as for \gsmax) then it has the same effect on the slope as \gsmax{} because there are many combinations of $D$ and $S ^ {0.5}$ that have same probability of colonization. If the probability of colonization is proportional to the product of *areal* stomatal size ($S$) and density ($\propto D S$ as for \fs) then it has the same effect on the slope as \fs{} because there are many combinations of $D$ and $S$ that have same probability of colonization. Alternatively, the probability of colonization may have a different scaling relationsip (neither 0.5 nor 1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we do not have theory to predict a stomatal size-density relationship that preserves the probability of colonization.

<!-- paragraph on what we don't know -->
In summary, the physical relationship between stomatal size, density, and conductance is well established [@harrison_influence_2019]. The same traits likely affect the probability of pathogen colonization, but we do not have a theoretical model that makes quantitative predictions. The inverse stomatal size-density relationship has usually been explained in terms of preserving stomatal conductance and/or stomatal cover, but selection by pathogens might alter scaling. To address these gaps, the goals of this study are to 1) introduce a spatially explicit model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs{}, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling.

@@

In this section, I introduce a spatially explicit model of pathogen colonization on a leaf surface. I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study. For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate. The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves (biotrophy). For example, motile bacterial cells can land and move around where fungi may germinate from a cyst and grow until they form an appresorium for infection. These very different biotropic movements on the leaf are treated identically here. I used Sympy version 1.3 [@meurer_sympy:_2017] for symbolic derivations.

@@

 Symbol | R      | Units  | Description
--------|--------|--------|-----------------------------------------------------
 $D,d$              | `D, d`    | $\textrm{mm}^{-2}$    | stomatal density ($d = \text{log}~D$)
 $f_s$              | `f_s`     | none                  | stomatal cover ($f_s = D S$)
 $g_\text{s,max}$   | `g_smax`  | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | anatomical maximum stomatal conductance 
 $g_\text{s,op}$    | `g_sop`   | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | operational stomatal conductance 
 $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface
 $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = 2 \pi R ^ 2$)
 $S,s$              | `S, s`    | $\mu\textrm{m}^2$     | stomatal size ($s = \text{log}~S$)
 $\theta_i$         | `theta_i` | radians               | angles between pathogen ($x_p,y_p$) and lines tangent to the circumfrence of stomate $i$
 $U$                | `U`       | $\mu\textrm{m}$       | interstomatal distance
 $v_i$              | `v_i`     | $\mu\textrm{m}$       | distance between pathogen ($x_p,y_p$) and stomate $i$
 $x_i,y_i$          | `x_i,y_i` | $\mu\textrm{m}$       | position of stomate $i$
 $x_p,y_p$          | `x_p,y_p` | $\mu\textrm{m}$       | starting position of pathogen

@@

Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015]. Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$. This assumption ignores veins, trichomes, and within-leaf variation in stomatal density. Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a). The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighbording stomata. The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$. By definition the stomatal density is the inverse of this area, such that $D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}$. Therefore, interstomatal distance can be derived from the stomatal density as:

@@

For example, if the density is $D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}$, then $U$ is `r round(sqrt(2 / (sqrt(3) * 1e-4)), 1)` $\mu \textrm{m}$. @parkhurst_diffusion_1994 described this result previously. I also make the simplifying assumption that stomata are perfectly circular with radius $R$. This may be approximately true for fully open stomata with kidney-shaped guard cells. Although I assume stomata are circular here, in calculating \gsmax, I assume typical allometric relationships between length, width, and pore area  (see [Supplementary Material]).

\begin{figure}
  \centering
	\includegraphics[width=6.5in]{../figures/fig1.jpeg}
	\caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circles represent idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the area. \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
	\label{fig:fig1}
\end{figure}

@@

Now imagine that a pathogen lands at a uniform random position within the focal region and must arrive at a stomate to colonize. If it lands on a stomate, then it infects the leaf with probability 1; if it lands between stomata, then it infects the leaf with probability $p_\text{locate}$. This is the probability that it locates a stomate, which I will derive below. The probabilities of landing on or between a stomate are \fs{} and $1 - f_S$, respectively. Hence, the total probability of colonization is:

@@

I assume that the pathogen cannot sense where stomata are and orients at random, thereafter traveling in that direction. If it successfully locates a stomate, it colonizes the leaf, but otherwise does not infect. If there is a high density of stomata and/or large stomata, the probability of locating a stomate increases. By assuming that stomata form an equilateral triangular grid (see above), we can extrapolate what happens in a focal triangle (Figure \ref{fig:fig1}b) by symmetry. Further, since an equilateral triangle can be broken up into six identical units (Figure \ref{fig:fig1}c), we can simply calculate $p_\text{locate}$ in this focal area. This implicitly assumes that the probability of colonzing stomata outside the focal area is 0 because they are too far away.

Imagine that the pathogens lands in position ($x_p,y_p$) within the triangle. The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}a). Therefore $x_c = U / 2$ and $y_c = \sqrt{3} U / 6$. The other stomata are at positions $(U/2, \sqrt{3} U /2)$ and $(U,0)$ (Figure \ref{fig:fig2}). $x_p$ and $y_p$ are defined as the horizontal and vertical distances, respectively, from the pathogen to the reference stomate at position $(0,0)$.

@@

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomata is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b.} Within the focal region, a pathogen can land within the stomate (white cirlce with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}

>>>>

---
title: "A stomatal model of anatomical tradeoffs between gas exchange and pathogen colonization"
runtitle: "Stomatal anatomy and pathogen colonization"
documentclass: frontiersSCNS
author:
  - name: Christopher D. Muir
    affiliation: '1'
    etal: Muir
    email: cdmuir@hawaii.edu
affiliation:
  - id: '1'
    department: School of Life Sciences
    institution: University of Hawaii
    city: Honolulu
    state: Hawaii
    country: USA
output: 
  rticles::frontiers_article:
    includes:
      in_header: header.tex
csl: frontiers.csl
bibliography: stomata-tradeoff.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

Stomatal pores control both leaf gas exchange and are one route for infection of internal plant tissues by many foliar pathogens, setting up the potential for tradeoffs between photosynthesis and defense. Anatomical shifts to lower stomatal density and/or size may also limit pathogen colonization, but such developmental changes could permanently reduce the gas exchange capacity for the life of the leaf. I developed and analyzed a spatially explicit model of pathogen colonization on the leaf as a function of stomatal size and density, anatomical traits which partially determine maximum rates of gas exchange. The model predicts greater stomatal size or density increases the probability of colonization, but the effect is most pronounced when the fraction of leaf surface covered by stomata is low. I also derived scaling relationships between stomatal size and density that preserves a given probability of colonization. These scaling relationships set up a potential anatomical conflict between limiting pathogen colonization and minimizing the fraction of leaf surface covered by stomata. Although a connection between gas exchange and pathogen defense has been suggested empirically, this is the first mathematical model connecting gas exchange and pathogen defense via stomatal anatomy. A limitation of the model is that it does not include variation in innate immunity and stomatal closure in response to pathogens. Nevertheless, the model makes predictions that can be tested with experiments and may explain variation in stomatal anatomy among plants. The model is generalizable to many types of pathogens, but lacks significant biological realism that may be needed for precise predictions.

@@

<!-- overview -->
Stomata evolved to regulate gas exchange in and out of the leaf [@hetherington_role_2003; @berry_stomata:_2010; @chater_origins_2017], but many foliar pathogens take advantage of these chinks in the leaf cuticular armor to infect prospective hosts [@zeng_plant_2010; @mclachlan_gate_2014; @melotto_stomatal_2017]. The stomatal and mesophyll conductance to CO$_2$ are two major limits to photosynthesis [@lawson_coordination_2018; @flexas_co2_2018] that are partially determined by stomatal anatomy. Since CO$_2$ conductance limits photosynthesis [@farquhar_stomatal_1982; @jones_partitioning_1985] and pathogen infection can reduce fitness [@gilbert_evolutionary_2002], this sets up a potential tradeoff between increased photosynthesis and defense against pathogens mediated by stomatal anatomy [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019]. For example, plants could increase photosynthetic rate by developing more stomata, but more stomata could result in more pathogen colonization. The optimal stomatal density, size, and arrangement on the leaf will depend on the fitness gains from increased gas exchange and fitness losses imposed by foliar pathogens, both of which depend on the environment. In the next two paragraphs I will review the relationship between stomatal anatomy, gas exchange, and foliar pathogen colonization. Then I will discuss why two anatomical traits, stomatal size and density, might be crucial components of a broader tradeoff between photosynthesis and pathogen defense.

<!-- anatomy and gas exchange -->
The stomatal density and maximum pore area set an anatomical upper limit on stomatal conductance [@brown_static_1900; @parlange_stomatal_1970; @franks_effect_2001; @franks_maximum_2009; @lehmann_effects_2015; @sack_developmental_2016; @harrison_influence_2019], but stomatal shape, distribution, and patterning also affect gas exchange. Smaller guard cells and dumbbell-shaped stomata of grasses can respond faster to environmental changes [@drake_smaller_2013], but responsiveness is further modulated by subsidiary cell anatomy and physiology [@franks_mechanical_2007; @raissig_mobile_2017; @gray_flanking_2020]. Stomatal clustering reduces gas exchange and photosynthesis because adjacent stomata interfere with one another [@dow_physiological_2014], diffusion shells overlap [@lehmann_effects_2015], and limitations on lateral diffusion of CO$_2$ in the mesophyll [@lawson_stomatal_2014 and references therein]. However, sparse clusters of small stomata could allow a leaf with low rates of gas exchange to have faster stomatal response compared to a leaf with large, low-density stomata [@papanatsiou_stomatal_2017]. Leaves with stomata on both lower and upper surfaces (amphistomatic or amphistomatous) supply more CO$_2$ to the mesophyll than hypostomatic (aka hypostomatous) leaves that only have stomata on the lower surface [@parkhurst_adaptive_1978; @gutschick_photosynthesis_1984; @parkhurst_intercellular_1990; @oguchi_leaf_2018]. In addition to anatomy, the pore area shrinks and expands in response to internal and external factors to regulate gas exchange dynamically [@buckley_how_2019]. For example, stomata typically open during the day and close at night in C$_3$/C$_4$ plants, but the opposite is true for CAM plants. Shade, high vapor pressure deficits, dry soil and other factors can cause stomata to (partially) close even in the middle of the day. Variation in how stomata respond to internal and external signals may explain as much of the variation in gas exchange across leaves as anatomy [@lawson_stomatal_2014].

<!-- anatomy and pathogens -->
Many types of foliar pathogens, including viruses [@murray_plant_2016], bacteria [@melotto_plant_2006; @underwood_role_2007], protists [@fawke_oomycete_2015], and fungi [@hoch_signaling_1987; @zeng_plant_2010] use stomatal pores to gain entry into the leaf. For example, rust fungi hyphae recognize the angle at which guard cells project from the leaf surface and use it as a cue for appressorium formation [@allen_appressorium_1991]. Oomycete pathogens can target open stomata on a leaf [@kiefer_host_2002]. Plants can limit colonization through innate immunity, called stomatal defense [recently reviewed in @melotto_stomatal_2017], by closing stomata after they recognize microbe-associated molecular patterns (MAMPs) on pathogen cells. Some bacterial pathogens have responded by evolving the ability to prevent stomatal closure, increasing their colonization of the leaf interior [@melotto_plant_2006]. In addition to stomatal closure, anatomical changes in stomatal density and/or size might provide another layer of defense against pathogen colonization. For example, infection increases in leaves with higher stomatal density [@mckown_association_2014; @tateda_host_2019; @dutton_bacterial_2019; @fetter_trade-offs_2019]. The positive effect of stomatal density on infection suggests that infection is limited by the number or size of locations for colonization, meaning that many individual pathogens must usually be unable to find stomata or other suitable locations for colonization. This is actually somewhat surprising given the ability of some pathogens to search for and sense stomata (see above).

<!-- Two paragraphs focusing the scope of this study -->
Stomatal anatomy could be a key link between gas exchange and pathogen colonization. Although many anatomical factors and stomatal movement affect gas exchange (see above), here I focus on the density and size of stomata in a hypostomatous leaf. Stomatal size refers to both the area of guard cells when fully open, from which one can calculate the pore area for gas exchange (see [Model]). For simplicity, I model a hypostomatous leaf, but consider the implications for amphistomatous leaves in the [Discussion]. Stomatal size and density not only determine the theoretical maximum stomatal conductance (\gsmax), but is also proportional to the operational stomatal conductance (\gsop) in many circumstances [@franks_plasticity_2009; @dow_integrated_2014; @franks_new_2014; @mcelwain_using_2016; @murray_consistent_2019]. \gsop{} is the actual stomatal conductance of plants in the field and is almost always below \gsmax{} because stomata are usually not fully open. Although they are not the same, the strong empirical relationship between \gsmax{} and \gsop{} means that anatomical \gsmax{} can be used as a proxy for \gsop{} without explicitly modeling dynamic changes in stomatal aperture (see [Discussion]). Stomatal size and density have also been measured on many more species than stomatal responsiveness, which may make it easier to test predictions.

After a pathogen reaches a host, it must survive on the leaf surface and colonize the interior [@beattie_secret_1995; @tucker_surface_2001]. For analytical tractability, I restrict the focus here to colonization by a pathogen using a random search on a leaf without stomatal defense (i.e. a leaf that cannot recognize pathogens and close stomata). Obviously, these simplifications ignore a lot of important plant-pathogen interaction biology. In the [Discussion], I delve further into these limitations and suggest future work to overcome these limitations. In order for pathogen-mediated selection on stomatal anatomy, I assume that the pathogen reduces host fitness once it colonizes [@gilbert_evolutionary_2002]. Susceptible hosts can lose much of their biomass or die, but even resistant hosts must allocate resources to defense or reduce photosynthesis because of defoliation, biotrophy, or necrosis around sites of infection [@bastiaans_ratio_1991; @mitchell_trophic_2003].

<!-- Purpose of study -->
The purpose of this study is to develop a theoretical framework to test whether variation in stomatal size and density arises from a tradeoff between gas exchange and pathogen colonization. Since stomatal size and density affect both gas exchange and pathogen colonization, selection to balance these competing demands could shape stomatal size-density scaling relationships. Botanists have long recognized that stomatal size and density are inversely correlated [@weiss_untersuchungen_1865; @ticha_photosynthetic_1982; @hetherington_role_2003; @sack_hydrology_2003; @franks_co_2009; @brodribb_unified_2013; @de_boer_optimal_2016], but the evolutionary origin of this relationship is not yet known. Here I argue that deleterious effects of pathogen infection could shape selection on this relationship. Explanations for inverse size-density scaling are usually cast in terms of preserving \gsmax{} and/or stomatal cover (\fs), defined at the fraction of epidermal area allocated to stomata [@de_boer_optimal_2016], because there are many combinations of stomatal size and density that have same \gsmax{} or same \fs{}:

@@

$D$ and $S$ are stomatal stomatal density and size, respectively (see Table 1 for a glossary of mathematical symbols and units). $b$ and $m$ are assumed to be biophysical and morphological constants, *sensu* @sack_developmental_2016 (see [Supplementary Material]). \fs{} is proportional to the more widely used stomatal pore area index [@sack_hydrology_2003; see [Supplementary Material]]. If size and density also affect pathogen colonization, then selection from foliar pathogens could significantly alter the size-density scaling relationship. The empirical size-density scaling relationship is linear on a log-log scale, determined by an intercept $\alpha$ and slope $\beta$:

@@

How would adding pathogens alter these predicted scaling relationships? For simplicity, consider two environments, one without foliar pathogens and one with lots. In the absence of foliar pathogens, we expect size-density scaling to preserve \gsmax{}, \fs{}, or some least-cost combination of them. What happens when we introduce pathogens? If stomatal size and density increase pathogen colonization, then selection will favor reduced size and/or density. This would change the intercept $\alpha$ but not the slope. The effect of foliar pathogens on the slope depends on the relationship between size, density, and probability of colonization. If the probability of colonization is proportional to the product of *linear* stomatal size ($S ^ {0.5}$) and density ($\propto D S ^ {0.5}$ as for \gsmax) then it has the same effect on the slope as \gsmax{} because there are many combinations of $D$ and $S ^ {0.5}$ that have same probability of colonization. If the probability of colonization is proportional to the product of *areal* stomatal size ($S$) and density ($\propto D S$ as for \fs) then it has the same effect on the slope as \fs{} because there are many combinations of $D$ and $S$ that have same probability of colonization. Alternatively, the probability of colonization may have a different scaling relationship (neither 0.5 nor 1) or may be nonlinear on a log-log scale. Unlike \gsmax{} and \fs{}, we do not have theory to predict a stomatal size-density relationship that preserves the probability of colonization.

<!-- paragraph on what we don't know -->
In summary, the physical relationship between stomatal size, density, and conductance is well established [@harrison_influence_2019]. Size and density also likely affect the probability of pathogen colonization, but we do not have a theoretical model that makes quantitative predictions. The inverse stomatal size-density relationship has usually been explained in terms of preserving stomatal conductance and/or stomatal cover, but selection by pathogens might alter scaling. To address these gaps, the goals of this study are to 1) introduce a spatially explicit model pathogen colonization on the leaf surface; 2) use the model to predict the relationship between \gsmax, \fs, and the probability of colonization; 3) work out what these relationships predict about stomatal size-density scaling. I analyzed an idealized, spatially explicit [Model] of how a pathogen lands on a leaf and finds a stomate to colonize the leaf using a random search. To my knowledge, this is the first model that makes quantitative predictions about the relationship between stomatal anatomy, the probability of colonization, and their impact on stomatal size-density scaling.

<!-- scraps -->
<!-- {By the same logic, if stomatal density were held constant, larger stomata should also increase colonization because they occupy more area, assuming stomatal responses are similar. One key difference between physiological and anatomical defenses is that stomatal closure reduces gas exchange transiently, whereas anatomy constrains gas exchange throughout the life of the leaf. -->

<!-- MOVE ELSEWHERE??: Although there have been many empirical studies on the effect that pathogens have on stomata, there is no theoretical framework in which to place these findings. Lack of a theoretical framework makes it difficult to answer general questions about how selection for pathogen defense constrains maximum rates of gas exchange. Although there is an empirical relationship between stomatal density and pathogen infection (see below), a theoretical framework based on anatomy may help explain empirical observations and generate new predictions. -->

<!-- Because we do not understand the relationship between stomatal anatomy and probability of colonization well, we cannot predict the relationship between \gsmax{} and infection.} -->

@@

In this section, I introduce a spatially explicit model of successful pathogen colonization on a leaf surface. I explain the model structure and assumptions here; the [Materials and Methods] section below describes how I analyzed the model to address the goals of the study. For generality, I refer to a generic ``pathogen'' that lands on leaf and moves to a stomate. The model is agnostic to the type of pathogen (virus, bacterium, fungus, etc.) and the specific biological details of how it moves. For example, motile bacterial cells can land and move around [@beattie_secret_1995] whereas fungi may germinate from a cyst and grow until they form an appresorium for infection [@tucker_surface_2001]. These very different tropic movements on the leaf are treated identically here. I do not model photosynthesis explicitly, but assume that stomatal conductance limits carbon fixation, even though the relationship is nonlinear. I used Sympy version 1.6.1 [@meurer_sympy:_2017] for symbolic derivations.

@@

 Symbol | R      | Units    | Description
--------|--------|----------|---------------------------------------------------
 $D$                | `D`       | $\textrm{mm}^{-2}$    | stomatal density
 $d$                | `d`       | $\textrm{mm}^{-2}$    | stomatal density (log-scale, $d = \text{log}~D$)
 \fs                | `f_s`     | none                  | stomatal cover ($f_s = D S$)
 $g_\text{s,max}$   | `g_smax`  | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | theoretical maximum stomatal conductance 
 $g_\text{s,op}$    | `g_sop`   | $\text{mol}~\text{m}^{-2}~\text{s}^{-1}$ | operational stomatal conductance 
 $H$                | `H`       | $\mu\textrm{m} ^{-1}$ | death rate of pathogen on leaf surface
 $R$                | `R`       | $\mu\textrm{m}$       | stomatal radius ($S = \pi R ^ 2$)
 $S$                | `S`       | $\mu\textrm{m}^2$     | stomatal size
 $s$                | `s`       | $\mu\textrm{m}^2$     | stomatal size (log-scale, $s = \text{log}~S$)
 $\theta_i$         | `theta_i` | radians               | angles between pathogen ($x_p,y_p$) and lines tangent to the circumference of stomate $i$
 $U$                | `U`       | $\mu\textrm{m}$       | interstomatal distance
 $v_i$              | `v_i`     | $\mu\textrm{m}$       | distance between pathogen ($x_p,y_p$) and stomate $i$
 $x_i,y_i$          | `x_i,y_i` | $\mu\textrm{m}$       | position of stomate $i$
 $x_p,y_p$          | `x_p,y_p` | $\mu\textrm{m}$       | starting position of pathogen

@@

Stomata develop relatively equal spacing to minimize resistance to lateral diffusion [@morison_lateral_2005], allow space between stomata [@dow_physiological_2014], and prevent stomatal interference [@lehmann_effects_2015]. Here I assume that stomata are arrayed in an equilateral triangular grid with a density $D$ and size (area) $S$ on the abaxial surface only, since most leaves are hypostomatoues [@muir_making_2015; but see [Discussion]]. This assumption ignores veins, trichomes, and within-leaf variation in stomatal density. Stomata are therefore arrayed in an evenly spaced grid (Figure \ref{fig:fig1}a). The interstomatal distance $U$, measured as the distance from the center of one stomata to the next, is the maximal diagonal of the hexagon in $\mu\textrm{m}$ that forms an equal area boundary between neighboring stomata. The area of a hexagon is $A_\textrm{hexagon} = \frac{\sqrt{3}}{2} U ^ 2$. By definition the stomatal density is the inverse of this area, such that $D = A^{-1}_\textrm{hexagon} = \frac{2}{\sqrt{3}} U ^ {-2}$. Therefore, interstomatal distance can be derived from the stomatal density as:

@@

For example, if the density is $D = 10 ^ 2~\textrm{mm}^{-2} = 10 ^ {-4}~\mu\textrm{m}^{-2}$, then $U$ is `r round(sqrt(2 / (sqrt(3) * 1e-4)), 1)` $\mu \textrm{m}$. @parkhurst_diffusion_1994 described this result previously. I also make the simplifying assumption that stomata are perfectly circular with radius $R$ when fully open. This may be approximately true for fully open stomata with kidney-shaped guard cells [@sack_developmental_2016 and references therein]. Although I assume stomata are circular here, in calculating \gsmax, I assume typical allometric relationships between length, width, and pore area  [@sack_developmental_2016; see [Supplementary Material]].

\begin{figure}
  \centering
	\includegraphics[width=6.5in]{../figures/fig1.jpeg}
	\caption{\textbf{A spatially explicit model of stomatal anatomy and pathogen colonization}. \textbf{a}. Stomata are assumed to be in a homogenous equilateral triangular grid, which means that we can extrapolate from \textbf{b.} a focal triangle to the entire leaf. The circles represent idealized stomata; the grey lines between them are for visualization. \textbf{c.} By symmetry, a single focal region within the focal triangle can be modeled and extrapolated to the rest of the triangle. \textbf{d.} The model assumes that a pathogen, depicted as a grey rod, lands somewhere on the leaf surface and will sucessfully locate a stomate if it moves at the correct angle, depicted by the grey polygons.}
	\label{fig:fig1}
\end{figure}

@@

Since stomata are arrayed in a homogeneous grid, we can focus on single focal triangle (Figure \ref{fig:fig1}b-c). Suppose that an individual pathogen (e.g. bacterial cell or fungal spore) lands at a uniform random position within the focal triangle and must arrive at a stomate to colonize. If it lands on a stomate, then it infects the leaf with probability 1; if it lands between stomata, then it infects the leaf with probability $p_\text{locate}$. This is the probability that it locates a stomate, which I will derive below. The probabilities of landing on or between a stomate are \fs{} and $1 - f_S$, respectively. Hence, the total probability of colonization is:

@@

I assume that the pathogen cannot sense where stomata are and orients at random, thereafter traveling in that direction. If it successfully locates a stomate, it colonizes the leaf, but otherwise does not infect. If there is a high density of stomata and/or large stomata, the probability of locating a stomate increases. By assuming that stomata form an equilateral triangular grid (see above), we can extrapolate what happens in the focal triangle (Figure \ref{fig:fig1}b) by symmetry. Further, since an equilateral triangle can be broken up into six identical units (Figure \ref{fig:fig1}c), we can simply calculate $p_\text{locate}$ in this focal area. This implicitly assumes that the probability of colonizing stomata outside the focal area is 0 because they are too far away. This assumption may be unrealistic for larger pathogens, such as fungi, whose hyphae can travel longer distances on the leaf surface [@perez-martin_tropic_2012]. In [Appendix 1: Spatially implicit model] I derive a simpler, but spatially *implicit* model that relaxes the assumption the pathogens must colonize a stomate within their focal triangle.

Consider a pathogen that lands in position ($x_p,y_p$) within the triangle. The centroid of the triangle is at position ($x_c,y_c$) and a reference stomate is at position $(0,0)$ (Figure \ref{fig:fig2}a). Therefore $x_c = U / 2$ and $y_c = \sqrt{3} U / 6$. The other stomata are at positions $(U/2, \sqrt{3} U /2)$ and $(U,0)$ (Figure \ref{fig:fig2}). $x_p$ and $y_p$ are defined as the horizontal and vertical distances, respectively, from the pathogen to the reference stomate at position $(0,0)$.

@@

\begin{figure}
\centering
\input{../figures/fig2.tex}
\caption{\textbf{Spatial representation of stomata and pathogen.} \textbf{a.} The pathogen starts at a uniform random position within the focal region denoted ($x_p,y_p$). Within the focal triangle, the reference stomate is at position (0,0) by definition, and other stomatal positions are determined by the interstomatal distance $U$. \textbf{b.} Within the focal region, a pathogen can land within the stomate (white circle with grey outline and radius $R$) or in the grey area. The outer borders of this area are shown and depend on $R$ and $U$. For a given position $x$, there is a minimum $y$-value ($y_\text{min}$, dashed line) and maximum $y$-value ($y_\text{max}$, solid line).}
\label{fig:fig2}
\end{figure}
